Formal Modeling and Verification of a Rate-Monotonic Scheduling Implementation with Real-Time Maude

### 2016-03-27 RMS/submission.tex
almost finished the submission

\subsection{Properties}
We consider two properties in this paper: schedulability and
correctness. By schedulability, we examine whether a given task set is
schedulable by the implementation. By correctness, we verify whether
the implementation schedules periodic tasks exactly with respect to
their priorities and periods.

To verify the schedulability of a given set of periodic tasks, we
define an atomic proposition \verb|taskTimeout| to hold if there
exists an \verb|error| appearing in $taskList$ of the current state,
that is, some task misses its deadline:
\begin{verbatim}
  op taskTimeout : -> Prop [ctor] .
  eq {L T STS HW ISRC} |= taskTimeout 
     = containError(L) .
\end{verbatim}
where \verb|containError| returns \verb|true| if there is an object
\verb|error| existing in \verb|L|. Then our desired
property--schedulability--can be formalized as the temporal logic
formula: \verb|[](~taskTimeout)|. As the property is not
\emph{clocked}, given an initial state \verb|init|, the following
untimed model checking command returns \verb|true| if the
schedulability property holds with no time limit; otherwise a trace
showing a counterexample is provided:
\begin{verbatim}
  (mc init |=u [](~taskTimeout) .)
\end{verbatim}

Another objective is to verify the correctness of the implementation.
The atomic proposition \verb|correct| is hence defined to hold if the
running periodic task is the one requested to be executed with the
highest priority:
\begin{verbatim}
  op correct : -> Prop [ctor] .
  ceq {L T STS HW ISRC} |= correct
      = if ID :: MaybeNat then shouldRun(ID, L)
        else true fi
      if ID := (HW).getPc .
\end{verbatim}
where \verb|shouldRun(ID, L)| returns \verb|true| if the task
identified by \verb|ID|, probably \verb|none|, is the one possessing
the highest priority among those whose status is \verb|READY|,
\verb|RUNNING| or \verb|INTERRUPT|. Note that in this paper, we do not
care the behaviors after some task misses its deadline. Therefore, the
correctness property is formalized by the temporal logic formula:
\verb|([]correct) \/ (correct U taskTimeout)|, and can be verified by
the following untimed model checking command provided an initial state
\verb|init|:
\begin{verbatim}
  (mc init |=u ([]correct) 
               \/ (correct U taskTimeout) .)
\end{verbatim}

@@

We have analyzed our model/implementation in ten different scenarios,
including both realistic ones provided by our industrial partner and
experimental ones designed by ourselves, four of them are described
below:
\begin{itemize}
\item Scenario~(i) with two tasks $\tau_1$ and $\tau_2$: $T_1=5ms$,
  $C_1=3ms$, $T_2=25ms$ and $C_2=7ms$.
\item Scenario~(ii) with two tasks $\tau_1$ and $\tau_2$: $T_1=5ms$,
  $C_1=2ms$, $T_2=25ms$ and $C_2=2.3ms$.
\item Scenario~(iii) with three tasks $\tau_1$, $\tau_2$ and $\tau_3$:
  $T_1=5ms$, $C_1=2.7ms$, $T_2=10ms$, $C_2=2ms$, $T_3=25ms$ and
  $C_3=3ms$.
\item Scenario~(iv) with three tasks $\tau_1$, $\tau_2$ and $\tau_3$:
  $T_1=5ms$, $C_1=2.5ms$, $T_2=10ms$, $C_2=1.5ms$, $T_3=15ms$ and
$C_3=4.5ms$.
\end{itemize}
Note that thanks to the powerful expressivity of Real-Time Maude, we
only need to define an initial state of sort \verb|System| to specify
a given task set. No necessity to modify the model is needed.

@@

Instantiating our model on dense time domain and choosing the
\emph{maximal time sampling strategy}, the results of the model
checking show that the correctness property holds in all scenarios. As
to the schedulability property, it holds in Scenarios~(i-iii), but
fails in Scenario~(iv).  One counterexample of the schedulability for
Scenario~(iv), returned by the model checking command, is pictured in
Figure~\ref{f:counterexample}, where the third task $\tau_3$ misses
its deadline at the time $15ms$.

@@

In general, completeness of untimed model checking cannot be achieved
for any systems, any time sampling strategies and any
properties. However, \"Olveczky and Meseguer proved the completeness
of untimed temporal logic model checking, under the maximal time
sampling strategy, for a large class of real-time
systems~\cite{DBLP:journals/entcs/OlveczkyM07a}:
\begin{theorem}[\cite{DBLP:journals/entcs/OlveczkyM07a}]
\label{t:completeness}
Given a \emph{time-robust} real-time rewrite theory $\cal R$, a set
$AP$ of \emph{tick-stabilizing} atomic propositions, an LTL formula
$\Phi$ (excluding the \emph{next} operator $\bigcirc$) whose atomic
propositions are contained in $AP$. The untimed temporal logic model
checking verifying $\Phi$ is \emph{complete} under the maximal time
sampling strategy.
\end{theorem}

Therefore, we achieve the following theorem, showing that the results
in Section~\ref{ss:results} are complete.
\begin{theorem}
\label{t:main}
Our approach using untimed model checking to verify the schedulability
and the correctness of our model is complete.
\end{theorem}
\begin{IEEEproof}
By showing that our model is time-robust and that the two defined
atomic propositions--\verb|taskTimeout| and \verb|correct|--are
tick-stabilizing, then using Theorem~\ref{t:completeness}.
\end{IEEEproof}

@@

Considering schedulability test, Liu and
Layland~\cite{DBLP:journals/jacm/LiuL73} gave the famous sufficient
condition that a set of periodic tasks is schedulable with respect to
RMS if $\displaystyle \Sigma^n_{i=1} C_i/T_i \le n(2^{1/n}-1)$
holds. Then a more sufficient condition, known as \emph{Hyperbolic
  Bound}, which has the same complexity as Liu and Layland's, was
proposed in~\cite{DBLP:journals/tc/BiniBB03}. On the other hand,
necessary and sufficient conditions for schedulability were derived
independently in~\cite{DBLP:journals/rts/SpruntSL89}
and~\cite{audsley1993deadline}, requiring more sophisticated analysis
on the task set. Nevertheless, all these results take no overhead into
account, being not as realistic as ours. Katcher et al. did consider
overhead in their schedulability analysis, under several models based
on different kinds of popular
implementations~\cite{DBLP:journals/tse/KatcherAS93}.  However, our
target implementation is not in their scope. Furthermore, compared
with those theoretical analyses, our approach based on formal modeling
and verification has three advantages. One is that if our
schedulability test answers ``no'', it returns at the same time a real
counterexample, which is able to guide our engineer to adjust the
design, by changing either the priorities or even the scheduling
algorithm. The second is that, when a fresh scheduling strategy is
applied, our analysis can be adjusted only by modifying the model,
while theoretical approaches may need thorough analyses and reasoning.
The last one is that, considering overhead and details of hardware do
introduce some kind of non-determinism into the model. For example, in
our model, if the running task is complete \emph{right} at the time
when an interrupt request is performed, two different behaviors are
possible: (i) the system performs task switching, during which the
interrupt request is masked, hence the scheduling may be delayed; (ii)
the system answers the interrupt request immediately, the switching
being delayed. Tackling such non-determinism via theoretical analyses
seems more complicated than our automatic approach.

\cite{TianD2011} and~\cite{DBLP:conf/iceccs/CuiDT14} also made use of
model checking to analyze the RMS algorithm along the same line but
with different languages and tools. The RMS algorithm is analyzed
using an extension of SPIN~\cite{DBLP:journals/tse/Holzmann97}
in~\cite{TianD2011}, while the logic programming language
TMSVL~\cite{DBLP:conf/icfem/HanDW12} and its model checker are applied
in~\cite{DBLP:conf/iceccs/CuiDT14}. The work presented
in~\cite{TianD2011} and~\cite{DBLP:conf/iceccs/CuiDT14} has two main
differences with ours. The first one, which is also the motivational
one, is that they considered only the ideal setting of the algorithm
as in~\cite{DBLP:journals/jacm/LiuL73}, instead of an implementation
which contains much more complex details. In fact, our model of the
implementation can easily degenerate to a model of the RMS algorithm,
if we let the times for scheduling and switching be zero. The other
difference is that when adding a new task into the task set, the
models in~\cite{TianD2011} and~\cite{DBLP:conf/iceccs/CuiDT14} should
be modified by explicitly defining a sub-model of the new task and its
behaviors. In particular, the scheduling part of the model
in~\cite{TianD2011} needs adjustments as well to include a new
task. Nevertheless, a new task set is specified in our approach merely
by giving a new initial state, without necessity to modify the model,
as already emphasized in Section~\ref{ss:results}. On the other hand,
\cite{TianD2011} used a discrete time domain in the model, while we
employ a generic time domain, which is flexible to be instantiated on
either discrete or dense ones. In~\cite{DBLP:conf/iceccs/CuiDT14}, the
time domain is dense, and a modeling strategy like our maximal time
sampling strategy is applied to reduce the state space. However, a
completeness statement like Theorem~\ref{t:main} was not given
in~\cite{DBLP:conf/iceccs/CuiDT14}.

@@

The following lemma is needed to prove the time-robustness of our
model:
\begin{lemma}[\cite{DBLP:journals/entcs/OlveczkyM07a}]
\label{l:timerobustness}
Let $\cal R$ be an object-oriented specification with a standard tick
rule, let the infinity element \verb|INF| be the only element in the
time domain which is not a normal time value, and let the time domain
be \emph{linear}. Then $\cal R$ is time-robust if the following
conditions are satisfied for all appropriate ground terms $t$ and
$r,r'$:
\begin{enumerate}
\item
[(i)] \verb|mte(delta(|$t$\verb|,|$r$\verb|))| $=$
\verb|mte(|$t$\verb|) monus |$r$, for all $r\le$
\verb|mte(|$t$\verb|)|;

@@

The following lemma is needed to prove the tick-invariance of our
defined propositions:
\begin{lemma}[\cite{DBLP:journals/entcs/OlveczkyM07a}]
\label{l:tickinv}
Let $\cal R$ be a time-robust object-oriented specification with a
standard tick rule, let the infinity element \verb|INF| be the only
element in the time domain which is not a normal time value, and let
the time domain be \emph{linear}. Then $\cal R$ is tick-invariant with
respect to a set $P$ of atomic propositions if \verb|{|$t$\verb|}|
$\simeq_P$ \verb|{delta(|$t$\verb|,|$r$\verb|)}| for all $t,r$ with
$r<$ \verb|mte(|$t$\verb|)|.
\end{lemma}

>>>>

\subsection{Properties}
We consider two properties in this paper: schedulability and
correctness. By schedulability, we examine whether a given task set is
schedulable by the implementation. By correctness, we verify whether
the implementation schedules periodic tasks exactly with respect to
the RMS algorithm.

To verify the schedulability of a given set of periodic tasks, we
define an atomic proposition \verb|taskTimeout| to hold if there
exists an \verb|error| in $taskList$ of the current state, that is,
some task misses its deadline:
\begin{verbatim}
  op taskTimeout : -> Prop [ctor] .
  eq {L T STS HW ISRC} |= taskTimeout 
     = containError(L) .
\end{verbatim}
where \verb|containError| returns \verb|true| if there is an
\verb|error| existing in \verb|L|. Then schedulability can be
formalized as the temporal logic formula: \verb|[](~taskTimeout)|. As
the property is not \emph{clock-related}, given an initial state
\verb|init|, the following untimed model checking command returns
\verb|true| if the schedulability property holds with no time limit;
otherwise a trace showing a counterexample is provided:
\begin{verbatim}
  (mc init |=u [](~taskTimeout) .)
\end{verbatim}

Another important objective is to verify the correctness of the
implementation.  The atomic proposition \verb|correct| is hence
defined to hold if the running periodic task is the one requested to
be executed with the highest priority:
\begin{verbatim}
  op correct : -> Prop [ctor] .
  ceq {L T STS HW ISRC} |= correct
      = if ID :: MaybeNat then shouldRun(ID, L)
        else true fi
      if ID := (HW).getPc .
\end{verbatim}
where \verb|shouldRun(ID, L)| returns \verb|true| if the task
identified by \verb|ID|, probably \verb|none|, is the one possessing
the highest priority among those whose status is not
\verb|DORMANT|. Note that during verification, we do not care the
behaviors after some task misses its deadline. Therefore, the
correctness property is formalized by the temporal logic formula:
\verb|([]correct) \/ (correct U taskTimeout)|, and can be verified by
the following untimed model checking command provided an initial state
\verb|init|:
\begin{verbatim}
  (mc init |=u ([]correct) 
               \/ (correct U taskTimeout) .)
\end{verbatim}

@@

We have analyzed our model in ten different scenarios, including both
realistic ones provided by our industrial partner and experimental
ones designed by ourselves, four of them are described below:
\begin{itemize}
\item Scenario~(i) with two tasks $\tau_1$ and $\tau_2$: $T_1=5ms$,
  $C_1=3ms$, $T_2=25ms$ and $C_2=7ms$.
\item Scenario~(ii) with two tasks $\tau_1$ and $\tau_2$: $T_1=5ms$,
  $C_1=2ms$, $T_2=25ms$ and $C_2=2.3ms$.
\item Scenario~(iii) with three tasks $\tau_1$, $\tau_2$ and $\tau_3$:
  $T_1=5ms$, $C_1=2.7ms$, $T_2=10ms$, $C_2=2ms$, $T_3=25ms$ and
  $C_3=3ms$.
\item Scenario~(iv) with three tasks $\tau_1$, $\tau_2$ and $\tau_3$:
  $T_1=5ms$, $C_1=2.5ms$, $T_2=10ms$, $C_2=1.5ms$, $T_3=15ms$ and
$C_3=4.5ms$.
\end{itemize}
Note that thanks to the powerful expressivity of Real-Time Maude, we
only need to define an initial state of sort \verb|System| to specify
a given task set. No necessity to modify the model is needed.

@@

Instantiating our model on dense time domain and choosing the
\emph{maximal time sampling strategy}, the results of the model
checking show that the correctness property holds in all scenarios. As
to the schedulability property, it holds in Scenarios~(i-iii), but
fails in Scenario~(iv).  One counterexample of the schedulability
within Scenario~(iv), returned by the model checking command, is
pictured in Figure~\ref{f:counterexample}, where the third task
$\tau_3$ misses its deadline at the time $15ms$.

We also randomly generate test scenarios under the above setting to
exam the efficiency of our approach for verifying the schedulability.
Among the $50$ generated test scenarios with $5$ tasks, we find that
the execution time of the model checking command for each scenario
varies from about $300ms$ up to a timeout, which is set to $90$
minutes. This is because the efficiency of model checking depends on
the scale of the state space. And the scale is further positively
correlated with $mn$, where $m$ is the upper bound of $timer$ and $n$
is the number of periodic tasks. By the generated test scenarios, it
turns out that the model checking command for schedulability in our
model is able to handle scenarios, where $mn$ is up to about $10^6$,
in an acceptable period of time say $90$ minutes.

@@

In general, completeness of untimed model checking cannot be achieved
for any systems, any time sampling strategies and any
properties. However, \"Olveczky and Meseguer proved the completeness
of untimed temporal logic model checking, under the maximal time
sampling strategy, for a large class of real-time systems possessing a
set of ``good'' properties that is called \emph{time-robustness}, and
for a set of ``good'' LTL formulae constructed by
\emph{tick-invariant} propositions\footnote{We avoid introducing the
  definitions of time-robustness and tick-invariance, due to the
  requirements of extra rewriting logic
  background.}~\cite{DBLP:journals/entcs/OlveczkyM07a}:
\begin{theorem}[\cite{DBLP:journals/entcs/OlveczkyM07a}]
\label{t:completeness}
Given a time-robust real-time rewrite theory $\cal R$, a set $AP$ of
tick-invariant atomic propositions, an LTL formula $\Phi$ (excluding
the \emph{next} operator $\bigcirc$) whose atomic propositions are
contained in $AP$. The untimed temporal logic model checking verifying
$\Phi$ is \emph{complete} under the maximal time sampling strategy.
\end{theorem}

Therefore, we achieve the following theorem, showing that the results
in Section~\ref{ss:results} are complete.
\begin{theorem}
\label{t:main}
Our approach using untimed model checking to verify the schedulability
and the correctness of our model is complete.
\end{theorem}
\begin{IEEEproof}
By showing that our model is time-robust and that the two defined
atomic propositions--\verb|taskTimeout| and \verb|correct|--are
tick-invariant, then using Theorem~\ref{t:completeness}. For a more
detailed proof, see the Appendix.
\end{IEEEproof}

@@

Considering schedulability test, Liu and
Layland~\cite{DBLP:journals/jacm/LiuL73} gave the famous sufficient
condition that a set of periodic tasks is schedulable with respect to
RMS if $\displaystyle \Sigma^n_{i=1} C_i/T_i \le n(2^{1/n}-1)$
holds. Then a more sufficient condition, known as \emph{Hyperbolic
  Bound}, which has the same complexity as Liu and Layland's, was
proposed in~\cite{DBLP:journals/tc/BiniBB03}. On the other hand,
necessary and sufficient conditions for schedulability were derived
independently in~\cite{DBLP:journals/rts/SpruntSL89}
and~\cite{audsley1993deadline}, requiring more sophisticated analysis
on the task set. Nevertheless, all these results take no overhead into
account, being not as realistic as ours. Katcher et al. did consider
overhead in their schedulability analysis, under several models based
on different kinds of popular
implementations~\cite{DBLP:journals/tse/KatcherAS93}.  However, our
target implementation is not in their scope. Furthermore, compared
with those theoretical analyses, our approach based on formal modeling
and verification has three advantages. One is that if our
schedulability test answers ``no'', it returns at the same time a real
counterexample, which is able to guide our engineer to adjust the
design, by changing either the priorities or even the scheduling
algorithm. The second is that, when a fresh scheduling strategy is
applied, our analysis can be adjusted only by modifying the model,
while theoretical approaches may need thorough analyses and reasoning.
The last one is that, considering overhead and details of hardware do
introduce some kind of non-determinism into the model. For example, in
our model, if the running task is complete \emph{right} at the time
when an interrupt request occurs, two different behaviors are
possible: (i) the system performs task switching, during which the
interrupt request is masked, hence the scheduling and initiation of
tasks may be delayed; (ii) the system answers the interrupt request
immediately, the switching being delayed. Tackling such
non-determinism via theoretical analyses seems more complicated than
our automatic approach.

\cite{TianD2011} and~\cite{DBLP:conf/iceccs/CuiDT14} also made use of
model checking to analyze the RMS algorithm along the same line but
with different languages and tools. The RMS algorithm is investigated
using an extension of SPIN~\cite{DBLP:journals/tse/Holzmann97}
in~\cite{TianD2011}, while the logic programming language
TMSVL~\cite{DBLP:conf/icfem/HanDW12} and its model checker are applied
in~\cite{DBLP:conf/iceccs/CuiDT14}. The work presented
in~\cite{TianD2011} and~\cite{DBLP:conf/iceccs/CuiDT14} has two main
differences with ours. The first one, which is also the motivational
one, is that they considered only the ideal setting of the algorithm
as in~\cite{DBLP:journals/jacm/LiuL73}, instead of an implementation
which contains much more complex details. In fact, our model of the
implementation can easily degenerate to a model of the RMS algorithm,
if we let the times for scheduling and switching be zero. The other
difference is that when adding a new task into the task set, the
models in~\cite{TianD2011} and~\cite{DBLP:conf/iceccs/CuiDT14} should
be modified by explicitly defining a sub-model of the new task and its
behaviors. In particular, the scheduling part of the model
in~\cite{TianD2011} needs adjustments as well to include a new
task. Nevertheless, a new task set is specified in our approach merely
by giving a new initial state, without necessity to modify the model,
as already emphasized in Section~\ref{ss:results}. On the other hand,
\cite{TianD2011} used a discrete time domain in the model, while we
employ a generic time domain, which is flexible to be instantiated on
either discrete or dense ones. In~\cite{DBLP:conf/iceccs/CuiDT14}, the
time domain is dense, and a modeling strategy like our maximal time
sampling strategy is applied to reduce the state space. However, a
completeness statement like Theorem~\ref{t:main} was not given
in~\cite{DBLP:conf/iceccs/CuiDT14}.

@@

Time-robustness is a set of properties expected from a well-behaved
real-time rewrite theory. We avoid introducing the accurate definition
of time-robustness, but instead give the following lemma to prove
time-robustness:
\begin{lemma}[\cite{DBLP:journals/entcs/OlveczkyM07a}]
\label{l:timerobustness}
Let $\cal R$ be an object-oriented specification with a standard tick
rule, and let the infinity element \verb|INF| be the only element in
the time domain which is not a normal time value.
%and let the time domain be \emph{linear}.
Then $\cal R$ is time-robust if the following conditions are satisfied
for all appropriate ground terms $t$ and $r,r'$:
\begin{enumerate}
\item
[(i)] \verb|mte(delta(|$t$\verb|,|$r$\verb|))| $=$
\verb|mte(|$t$\verb|) monus |$r$, for all $r\le$
\verb|mte(|$t$\verb|)|;

@@

The following lemma is needed to prove the tick-invariance of our
defined propositions:
\begin{lemma}[\cite{DBLP:journals/entcs/OlveczkyM07a}]
\label{l:tickinv}
Let $\cal R$ be a time-robust object-oriented specification with a
standard tick rule, and let the infinity element \verb|INF| be the
only element in the time domain which is not a normal time value.
%and let the time domain be \emph{linear}. 
Then $\cal R$ is tick-invariant with respect to a set $P$ of atomic
propositions if \verb|{|$t$\verb|}| $\simeq_P$
\verb|{delta(|$t$\verb|,|$r$\verb|)}| for all $t,r$ with $r<$
\verb|mte(|$t$\verb|)|.
\end{lemma}
